---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---

## Exploring the NOAA Storm Database: Analyzing the impact of severe weather events on population health and the economy from 1950 to 2011.

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and local government institutions. Those severe events can result in fatalities, injuries, property and other damages to the economy. Being able estimate the impact of those events is a major area of focus to define public policy and investments in prevention.

This analysis will focus on finding answer to two major questions:

1. Which types of events are most harmful with respect to population health
2. Which types of events have the greatest economic consequences

across the United States.

This project uses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database that recorded events from the year 1950 until November 2011. This database tracks characteristics of major storms and other significant weather events that have sufficient intensity to cause deaths, injuries, damages to properties and other damages to economic activities.

The data file and supporting documentation is available for download from the following sources:

1. [The main NOAA database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

2. [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

3. [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


## Data Processing

### 1. Setting up the working environment, downloading and loading data
To start we load our libraries

```{r setup}
require(ggplot2)

require(reshape2)
```

and set the working directory
```{r set_wd}
pwd <- getwd() 
if (!is.null(pwd)) { 
                setwd(pwd) 
        } else { 
                print("Directory not found") 
        }

```

Dowloading, saving and loading the data that will be used to conduct the analysis

```{r download_file}
if(!file.exists("stormData.csv.bz2")) {
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
        destfile = "stormData.csv.bz2")
}
```

Loading the file onto R
```{r load_file}
data <- read.csv(bzfile("stormData.csv.bz2"), sep = ",", header = TRUE)
```

### 2. Analyzing the data structure
```{r data_structure}
#Here we are looking for the names of the different variables
names(data)

# Now we look at the general size of the data set
dim(data)

# And more specifically to the structure of the variables
str(data)
```


```{r}

# Change parameter names to lowercase.
colnames(data) <- tolower(colnames(data))
```

Our analysis focuses on identifying which events caused most economic and health damages. Therefore, the dataset is filtered to keep only the columns containing relevant data i.e. "evtype", "fatalities", "injuries", "propdmg", "propdmgexp", "cropdmg", "cropdmgexp".  
    
```{r}
# Subset the data set to include only the parameters of interest for the analysis.

data <- subset(x = data, 
                   subset = (evtype != "?" & 
                            (injuries > 0 | fatalities > 0 | propdmg > 0 | cropdmg > 0)),
                   select = c("evtype", 
                            "fatalities", 
                            "injuries", 
                            "propdmg", 
                            "propdmgexp", 
                            "cropdmg", 
                            "cropdmgexp"))
```


```{r}
# We change all damage exponents to uppercase 
data$propdmgexp <- toupper(data$propdmgexp)
data$cropdmgexp <- toupper(data$cropdmgexp)

```

We then apply a conversion that changes the letter to the relevant numeric value for the Property damages.
The actual amounts specified in the DMGEXP columns are:
- B for Billion
- M for Million
- K for Thousand

```{r}
propDmg <-  c("\"\"" = 10^0,
                "-" = 10^0, 
                "+" = 10^0,
                "0" = 10^0,
                "1" = 10^1,
                "2" = 10^2,
                "3" = 10^3,
                "4" = 10^4,
                "5" = 10^5,
                "6" = 10^6,
                "7" = 10^7,
                "8" = 10^8,
                "9" = 10^9,  
                "H" = 10^2,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)

# Providing the correct formatting of data for the analysis
data$propdmgexp <- propDmg[as.character(data$propdmgexp)]
data$propdmgexp[is.na(data$propdmgexp)] <- 10^0
```

Applying the conversion to the Crop damages as well.

```{r}
cropDmg <-  c("\"\"" = 10^0,
              "?" = 10^0, 
              "0" = 10^0,
              "K" = 10^3,
              "M" = 10^6,
              "B" = 10^9)

# Providing the correct formatting of data for the analysis
data$cropdmgexp <- cropDmg[as.character(data$cropdmgexp)]
data$cropdmgexp[is.na(data$cropdmgexp)] <- 10^0
```

### 3.1 Processing the data to be used in the analysis of health consequences

```{r health_agg}
# Aggregating number of fatalities and injuries per evtype into healthData dataframe
healthData <- aggregate(cbind(fatalities, injuries) ~ evtype, data=data, FUN=sum)

# Adding fatalities and injuries
healthData$tot <- healthData$fatalities + healthData$injuries
```

```{r health_clean_sort}
# Removing rows with zero values
healthData <- healthData[healthData$tot > 0, ]

# Sorting data in descending order
healthData <- healthData[order(healthData$tot, decreasing=TRUE), ]

# Filtering the dataframe to show the top 10 impact event types
healthDataTop <- healthData[1:10, ]

# Removing the column for totals
healthDataTop$tot <- NULL
```

```{r health_melt}
# Changing the data set format from wide to long for graphing
healthDataTopMelt <- melt(healthDataTop, id.vars="evtype")
```

### 3.2 Processing the data to be used in analyzing the economic consequences

```{r econ_agg}
# Combining propdmg and propdmgexp parameters into a single parameter "propertyloss"
data$proploss <- data$propdmg * data$propdmgexp

# Combining cropdmg and cropdmgexp parameters into a single parameter "croploss"
data$croploss <- data$cropdmg * data$cropdmgexp

# Aggregating amount of proploss and croploss per evtype into economicData dataframe
econData <- aggregate(cbind(proploss, croploss) ~ evtype, data=data, FUN=sum)

# Adding total loss column to economicData
econData$tot <- econData$proploss + econData$croploss
```

```{r econ_clean_sort}
# Removing rows with zero values
econData <- econData[econData$tot > 0, ]

# Sorting the economy data in descending order
econData <- econData[order(econData$tot, decreasing=TRUE), ]

# Creating a dataframe of highest economy impacting event types
econDataTop <- econData[1:10, ]

# Removing the column for totals
econDataTop$tot <- NULL
```

```{r econ_melt}
# Changing the data set format from wide to long for graphing
econDataTopMelt <- melt(econDataTop, id.vars="evtype")
```

## Results Analysis

In this first chart we analyse the impact that adverse weather event have on population health.  


```{r}
# Creating the chart for health injuries
healthChart <- ggplot(healthDataTopMelt, aes(x = reorder(evtype, -value), y = value)) +
                        geom_bar(stat = "identity", aes(fill = variable)) +
                        scale_y_sqrt("Event Frequency") +
                        xlab("Event Type") +
                        theme(axis.text.x = element_text(angle=45, hjust = 1)) +
                        ggtitle("Health Impact of Top 10 US Weather Events")

# Displaying the chart
print(healthChart)
```


From the chart it is clear that tornadoes have the highest health impact in terms of both fatalities and injuries. 
Generally, we can see that the number of fatalities is low compared to the number of injuries for all of the weather events.  



In this second chart we analyse the economic impact of these adverse weather.  


```{r}
# Creating the chart for economic damages
econChart <- ggplot(econDataTopMelt, aes(x = reorder(evtype, -value), y = value)) + 
        geom_bar(stat = "identity", aes(fill = variable)) + 
        scale_y_sqrt("Damage ($)") + 
        xlab("Event Type") + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
        ggtitle("Economic Impact of Top 10 US Weather Events")

# Displaying the chart
print(econChart)
```

From the chart we can see that it is clear that tornadoes have the highest health impact in terms of both fatalities and injuries. 
Tornadoes, as in the previous chart, are responsible for the majority of economic damage almost all of which is attributable to property losses.
The chart also shows that the biggest contributor to crop loss in economic terms is drought which is responsible for about three million dollars in damages.
